---
title: "filter"
author: "Onkar Singh"
date: "10/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
```

## source data
### [video & data] (https://www.youtube.com/watch?v=JuTdj2j-9Kg)


Two ways to reading/coverting the date columns from excel to R

The base function as.Date handles this type of conversion nicely. The trick is that you have to provide the origin, which in Excel is December 30, 1899. 

Read further for other options: 
https://stackoverflow.com/questions/43230470/how-to-convert-excel-date-format-to-proper-date-in-r/43230524#43230524



```{r}
data <- readxl::read_xlsx("filter-examples.xlsx", sheet = "data") %>% 
   mutate(Date_Joined = as.Date(Date_Joined, origin = "1899-12-30")) 
  # janitor::clean_names() %>% 
  # mutate(date_joined = lubridate::as_date(date_joined)) #This is the best way but commenting it out because we want the Column Name in Camel Case so I have changed the Date Joined column name in excel to Date_Joined by filling the space with a underscore for R to read it naturally n not throw any error
  


data
```

### Filter rows where gender == female

```{r}
data %>% 
    filter(Gender == "Female")
```


### Filter rows where Gender is female and only print the name column


```{r}
data %>% 
  filter(Gender == "Female") %>% 
  select(Name)
```

### Filter dynamically by assigning the filter criteria to a variable

```{r}
gender_filter <- "Female"

data %>% 
  filter(Gender == gender_filter) %>% 
  select(Name)
```


## Count the number of values after filtering

```{r}
data %>% 
  filter(Gender == gender_filter) %>% 
  select(Name) %>% 
  count()
```


### We want to filter all the male employees in the website department

```{r}
data %>% 
  filter(Gender == "Male", 
         Department == "Website")
```

### filter staff from a particular department with salary above 90K: AND Condition Example

```{r}
data %>% 
    filter(Department == "Website", 
         Salary > 90000)
```

### filter staff from a particular department or with salary above 90K: OR Condition Example
outputs everybody in the Website or have a salary over 90K
```{r}
data %>% 
    filter(Department == "Website" |
         Salary > 90000)
```

outputs everybody in the Sales or have a salary over 80K

```{r}
data %>% 
    filter(Department == "Sales" |
         Salary > 80000)
```


### Anybody who joined after this particular date or who have a salary less than 65K

```{r}
data %>% 
    filter(Date_Joined > "2019-01-01" |
         Salary < 65000)
```

### Not Condition: Print Any Cells where the geography is not USA from 2nd data table

```{r}
data2 <- readxl::read_excel("filter-examples.xlsx", sheet = "data2")

data2 %>% 
  filter(Geography != "USA")
```

### Not Condition: Print Any Cells where the geography is not USA and only print the sales column

```{r}
data2 %>% 
  filter(Geography != "USA") %>% 
  select(Amount) %>% 
  count(name = "total_row_count", sum = sum(Amount)) %>% 
  select(total_row_count, sum)
```



### Complex Condition 
Geography is either USA or UK and the amount is grater than 2000 dollars


```{r}
data2 %>% 
  filter(Geography %in% c("USA", "UK"), 
         Amount> 2000) %>% 
  count(name = "total_row_count", Total_Amount = sum(Amount), Total_Units = sum(Units)) %>% 
  select(total_row_count, everything())
  
```


### What if there is an error in the formula or say the value for which we are flitering doesnt existing in R it will simply return 0 rows but excel throws error

```{r}
data2 %>% 
  filter(Geography %in% c("USA", "UK"), 
         Amount> 50000) %>% # we know that there is no amount greater than 50000 in the table
  count(name = "total_row_count", Total_Amount = sum(Amount), Total_Units = sum(Units)) %>% 
  select(total_row_count, everything())
  
```

 ## Filter Problems
 
### 1. Print names of all people in either Sales or Website department

```{r}
data %>% 
  filter(Department %in% c("Sales", "Website")) %>% 
  select(Name)
```

### 2. How many people have joined in the first 3 months of 2018 & 2019?

```{r}
data %>% 
  filter (Date_Joined >= "2018-01-31", 
          Date_Joined <= "2018-03-31" |
            Date_Joined >= "2019-01-31", 
          Date_Joined <= "2019-03-31")


```

### 3. Print data for all 50% dark bites sold in New Zealand

```{r}
data2 %>% 
  filter(Geography == "New Zealand", 
         Product == "50% Dark Bites") 
  
```

### 4. What is the total amount where country is neither India nor Australia

```{r}
data2 %>% 
  filter(!Geography %in% c("India", "Australia")) %>% 
  count(name = "total_rows_count", sum = sum(Amount)) %>% 
  select(total_rows_count, sum)
```




### 5. Print data for all sales from other people where the amount is > average sales of Gunar Cockshoot

```{r}
data2 %>% 
  filter(`Sales Person` == "Gunar Cockshoot") %>% 
  count(avg_GC = mean(Amount), name = "total_row_count") %>% 
  select(total_row_count, avg_GC)
```

```{r}
data2 %>% 
  filter(`Sales Person` != "Gunar Cockshoot", 
Amount > 3395)
```


## Advanced Filter Trick Example by Leila Gharani

[link of the video n file](https://www.youtube.com/watch?v=evrnIuDRtsQ)


```{r}
data3 <- readxl::read_excel("Advanced_Filter_Trick_Example.xlsx", sheet = "Data") %>% 
  mutate(Date = lubridate:: as_date(Date))

data3
```


### Filter with a date greater than a value plus string match

```{r}
data3 %>% 
  filter(Date > "2019-07-01", 
         str_detect(`Article Description`, "Lap"))
```

