---
title: "onpenxlsx_creating_spreadsheet_from_r"
output: html_document
date: "2022-11-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[Tweet from Neil Currie](https://twitter.com/neilgcurrie/status/1596186976717193216)
```{r cars}
library(openxlsx)
library(dplyr)
library(ggplot2)
library(glue)
```

```{r}
#install.packages(c('openxlsx', 'dplyr', 'ggplot2', 'glue'))
# install.packages('here')
```

```{r}
# Make some dummy data

diamonds_xl <- diamonds |> 
  group_by(cut) |> 
  summarize (n = n(), 
             average_price = mean(price), 
             total_price = sum(price))

# Create workbook
wb <-  createWorkbook()

addWorksheet(wb, "diamonds")
addWorksheet(wb, "starwars")

filename <- glue("{here::here()}/data/workbook.xlsx")

saveWorkbook(wb, filename)
  
```

## 2. Adding data

Adding data is straightforward too.

You can specify details like what sheet to use, which column and row to start writing, and if you should include column and row names or numbers.

Data is written in blocks starting in the row and column you specify.
```{r}
writeData(wb, sheet = 'diamonds', x = diamonds_xl, startCol = 2, startRow = 1)

writeData(wb, sheet = 'starwars', x = starwars, rowNames = TRUE)
```

## 3. Adding formulas

Spreadsheets are driven by formulas. You can do all sorts of calculations in R but if you want your spreadsheet to react to user input you need a formula.

Formulas can be written as strings and added using the writeFormula function.

We will add in a Target Price and Price Difference columns. The Price Difference column will be formula based.

The user will be able to write whatever they want in the Target Price cells and the formula will update.

```{r}
## Add headings
writeData(wb, sheet = 'diamonds', x = 'Target Price', startCol = 7)
writeData(wb, sheet = 'diamonds', x = 'Price Difference', startCol = 8)

# The user will write a target price manually later
# add a formula to calculate the difference 
# you don't need the equal sign

price_diff <- glue("G{2:6}-E{2:6}")

writeFormula(wb, sheet = 'diamonds', x = price_diff, startCol = 8, startRow = 2)
```


## 4. Formatting cells

Now we have a functional spreadsheet. But it looks pretty basic.

To make it look better and be more usable we need to style it.

First we need to create the styles. We then add that style to whatever cells we want.

```{r}
# Some different styles
style_header = createStyle(fontSize = 16, fontColour = 'blue', border = 'bottom', wrapText = TRUE)

style_currency = createStyle(numFmt = "CURRENCY")

style_user_input = createStyle(fgFill = 'yellow')

# Write
addStyle(wb, sheet = 'diamonds', style_header, rows = 1, cols = 2:8, gridExpand = TRUE)
addStyle(wb, sheet = 'diamonds', style_currency, rows = 2:6, cols = 4:5, gridExpand = TRUE)
addStyle(wb, sheet = 'diamonds', style_user_input, rows = 2:6, cols = 7, gridExpand = TRUE)
```

## 5. Formatting the spreadsheet

We can't really see the data in the header. And it is a bit messy. The code attached will tidy that up.

Freezing panes can also be useful when data is bigger than the screen.

```{r}
setColWidths(wb, sheet = 'diamonds', cols = c(1, 6), widths = 3)
setColWidths(wb, sheet = 'diamonds', cols = c(2:5, 7:8), widths = 15)
setRowHeights(wb, sheet = 'diamonds', rows = 1, heights = 50)

freezePane(wb, sheet = "diamonds", firstActiveCol = 3)

colnames(diamonds_xl) <- c("Cut", "Num Diamonds", "Average Price", "Total Price")

writeData(wb, sheet = "diamonds", x = diamonds_xl, startCol = 2, startRow = 1)
```

## 6. Conditional formatting

Conditional formatting is useful in all sorts of situations. A common one is highlighting if values are positive or negative so they stand out to the user.

And doing this is similar to adding a regular style.

```{r}
neg_style <- createStyle(fontColour = "#9C0006", bgFill = "#FFC7CE")
pos_style <- createStyle(fontColour = "#006100", bgFill = "#C6EFCE")

conditionalFormatting(wb, sheet = "diamonds", cols = 8, rows = 2:6, rule = "<0", style = neg_style)
conditionalFormatting(wb, sheet = "diamonds", cols = 8, rows = 2:6, rule = ">0", style = pos_style)

saveWorkbook(wb, filename, overwrite = TRUE)
```

To recap:

You can create spreadsheets from scratch directly from R with openxlsx.

You can add data, formulas, format cells and freeze panes plus more.

This is ideal for automating processes, saving time and improving quality.

Great thread! Just to add one more {openxlsx} function that I find quite helpful: 

`writeDataTable()` writes and formats the data as a filterable Excel table. It lets you even specify which table template to use.

There's also the rhandsontable package which is really handy for building spreadsheets into shiny apps

